Fluent statistical computing interfaces for biological data analysis

### 2020-07-16 Rmd/00-e-preface.Rmd
add preface and rejig intro

>>>>

Chapter \@ref(ch:plyranges) has been published in _Genome Biology_. It
won the ACEMS Business Analytics Prize for Best Paper in 2019. Chapter \@ref(ch:fluentGenomics) has been published in _F1000 Research_.  Chapter
\@ref(ch:intron) is based on my software contributions to the submitted
article @Lee2019-mf. Chapter \@ref(ch:embedding) has not been submitted yet.

## Open Research

This thesis has been written using R Markdown with the **bookdown**
package [@Xie2016-oa; @Xie2017-ke]. All materials required to build the
thesis from source can be obtained from https://github.com/sa-lee/thesis. An
online version of the thesis is available at https://sa-lee.github.io/thesis.

### 2020-07-16 Rmd/01-intro.Rmd
add preface and rejig intro

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. It is organised in 
according to the aforementioned workflow steps. To begin we discuss wrangling
biological data using a coherent representation and programming interface; 
section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). Next, we look at integrating data and 
model outputs over genomic regions to gain biological insight; \@ref(sec:ir) 
introduces a framework for incorporating and discovering 'interesting' genomic
regions, described further in chapters \@ref(ch:fluentGenomics) and 
\@ref(ch:intron). Third, we consider the challenges in visualising  
high dimensional data; section \@ref(sec:va) introduces a visual analytics 
approach for understanding non-linear dimension reduction techniques described
in chapter \@ref(ch:tsne). Lastly, in chapter 
\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
developments.

@@

Chapter \@ref(ch:plyranges) shows that the *tidy data* semantic is applicable
to range-based genomics data and develops a *fluent* interface to transforming
it called *plyranges*. The software provides a framework to an assist
an analyst to compose queries on genomics datasets. This chapter has been 
published in @Lee2019.

## Integration and representation of genomic data structures {#sec:ir}

It is rare that a biological data analysis will involve
a single measurement assay or that only one aspect of a measurement assay
will be of interest to the biological question under study. While there are 
many approaches to integrating data sets from multiple assays using multivariate
statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
we describe a simple end-to-end workflow for integrating 
results along the genome using *plyranges*. This workflow also shows that
our grammar based approach does not impair interoperability between
the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
together. This chapter has been published in @Lee2020-ie.

Moreover, in chapter \@ref(ch:intron) we explore the limits of the *tidy data*
semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
data by developing a new software tool called *superintronic*. We show that the
long-form tidy representation is an effective way of combining the experimental
design and reference annotations into a single genomic data structure for
exploration. We use *superintronic* to develop a framework for discovering
interesting regions of coverage and apply our approach to integrating intron
signal from RNA-seq data. This chapter is based on the preprint @Lee2019-mf.

## Visual analytics for dimension reduction workflows {#sec:va}

In the final chapter, we move away from data wrangling and towards the 
integration of visualisation with model-based summaries of high-dimensional
data sets. We focus on a common tool for EDA (especially applied to single-cell
transcriptomics): dimension reduction (DR). We consider the incorporation of 
interactive and dynamic graphics to assist analysts in using DR techniques
for cluster orientation tasks. In particular, we advocate for the use of tours
[@Cook1995-bi] alongside a DR visualisation to highlight potential pitfalls and 
distortions of obtained from a DR method. This approach acknowledges that
there is no 'one' best visualisation or DR for a high-dimensional dataset,
and we often want to have an understanding of both the global and local structure
within our data.

Chapter \@ref(ch:tsne) introduces
a software package called *liminal* for constructing these views and
a user interaction framework for identifying distortions. We present several
case studies using data sets from both single-cell transcriptomics and 
machine learning  for using our approach to diagnose the quality of 
results obtained via popular non-linear DR methods
like t-distributed stochastic neighbour embeddings (t-SNE).

>>>>

This thesis focuses on core aspects of EDA as part of a biological data
science workflow: wrangling, integration and visualisation, with a focus 
on applications to genomics and transcriptomics. To begin we discuss wrangling
biological data using a coherent representation and programming interface; 
section \@ref(sec:tidygranges)
introduces a grammar-based framework for transforming genomics data that is 
described in chapter \@ref(ch:plyranges). We then look at integrating data and 
model outputs over genomic regions to gain biological insight; \@ref(sec:intron) 
introduces a framework for incorporating genomic regions over multiple assays, 
described in chapter \@ref(ch:fluentGenomics), while \@ref(sec:coverage)
discusses finding 'interesting' genomic regions via multiple summaries of a 
single assay, described in \@ref(ch:intron). Next, we consider the challenges in 
visualising  high dimensional data; section \@ref(sec:va) introduces an 
interactive visualisation approach for understanding non-linear dimension 
reduction  techniques described in chapter \@ref(ch:tsne). Lastly, in chapter 
\@ref(ch:conclusion) I describe the outputs of my thesis and plans for future 
developments.

@@

Chapter \@ref(ch:plyranges) shows that the *tidy data* semantic is applicable
to range-based genomics data and develops a *fluent* interface to transforming
it called *plyranges*. The software provides a framework to an assist
an analyst to compose queries on genomics datasets. This chapter has been 
published as @Lee2019.

## Integration of genomic data structures {#sec:integrate}

It is rare that a biological data analysis will involve
a single measurement assay or that only one aspect of a measurement assay
will be of interest to the biological question under study. While there are 
many approaches to integrating data sets from multiple assays using multivariate
statistical techniques [@Meng2016-pz], in chapter \@ref(ch:fluentGenomics)
we describe a simple end-to-end workflow for integrating 
results along the genome using *plyranges*. This workflow also shows that
our grammar based approach does not impair interoperability between
the *tidyverse* and *Bioconductor* approaches, and in fact they work seamlessly
together. This chapter has been published as @Lee2020-ie.

## Representation of genomic data structures{#sec:coverage}

In chapter \@ref(ch:intron) we explore the limits of the *tidy data*
semantic by extending *plyranges* to analyse coverage estimated on RNA-seq
data by developing a new software tool called *superintronic*. We show that the
long-form tidy representation is an effective way of combining the experimental
design and reference annotations into a single genomic data structure for
exploration. We use *superintronic* to develop a framework for discovering
interesting regions of coverage and apply our approach to integrating intron
signal from RNA-seq data. This chapter is based on my software and analysis
contributions to the preprint @Lee2019-mf.

## Interactive visualisation for high-dimensional data {#sec:va}

In the final chapter, we move away from data wrangling and towards the 
integration of visualisation with model-based summaries of high-dimensional
data sets. We focus on a common tool for EDA (especially applied to single-cell
transcriptomics): non-linear dimension reduction (NLDR). We consider the incorporation of 
interactive and dynamic graphics to assist analysts in using NLDR techniques
for cluster orientation tasks. In particular, we advocate for the use of tours
[@Cook1995-bi] alongside an NLDR visualisation to highlight potential pitfalls 
and distortions of obtained from an NLDR method. This approach acknowledges that
there is no 'one' best visualisation or dimension reduction for a 
high-dimensional dataset, and we often want to have an understanding of both the global and local structure within our data.

Chapter \@ref(ch:tsne) introduces a software package called *liminal* for 
constructing these views and a user interaction framework for identifying 
distortions. We present several case studies using data sets from both 
single-cell transcriptomics and  machine learning  for using our approach to diagnose the quality of  results obtained via popular NLDR methods
like t-distributed stochastic neighbour embeddings (t-SNE).
